### 2025-07-03 lecturenotes_aima104/ch4_aima104.tex
orthogonal

@@

\subsection{Orthogonal Sets}
A set of vectors $\{ u_1, ...u_p\}$ in a vector space $V$ is said to be \textbf{orthogonal set} if each pair of distinct vectors from the set is orthogonal,i.e for all $u_i, u_j \in V$ we have $u_i.u_j = 0$ whenever $i \neq j$.

@@

\subsection{Orthogonal Projection}
The coordinate $c_j$ of $y$ in Theorem~\ref{orthopro} is actually orthogonal projection of $y$ into the vector $u_j$. This can be generalized. For any given vector $u$. The orthogonal projection of a vector $y$ on $u$ is given by the formula $\displaystyle \hat{y} = \frac{y.u}{u.u}$

>>>>

\subsection{Exercise}
\begin{enumerate}
\item Compute the following for the given vectors: \\
  $u=
  \begin{bmatrix}
    -1 \\ 2
  \end{bmatrix}, \vspace{4mm}
   v=
   \begin{bmatrix}
     4 \\6
   \end{bmatrix},
   vspace{5mm}
   w=
   \begin{bmatrix}
     3 \\ -1 \\ 5
   \end{bmatrix},\vspace{4mm}
   x=
   \begin{bmatrix}
     6 \\ -2 \\3
   \end{bmatrix}
   $
   \begin{multicols}{2}
     a). $\displaystyle u.u, \;\; v.u, \;\; \frac{v.u}{u.u}, \;\; \Vert v \Vert $
     \columnbreak

     b). $\displaystyle w.w, \;\; x.w, \;\; \frac{x.w}{w.w}, \;\; \Vert x \Vert$
   \end{multicols}
 \item Find the distance between $u$ and $v$, and $w$ and $x$.
 \item Use matrix product and transpose definition to verify, property-2 and 3 of the inner product.
 \item Explain why $u.u \geq 0$. When is $u.u =0$.?
\end{enumerate}

@@

\subsection{Exercise}
\begin{enumerate}
 \item Verify parallelogram law: $\Vert u+v \Vert ^2 = 2 \Vert u \Vert ^2 + 2 \Vert v \Vert ^2$.
 \item Suppose $y$ is orthogonal to $u$ and $v$. Show that $y$ is orthogonal to every $w$ in $Span\{u,v\}$.
 \item Let $W$ be a subspace of $\mathbb{R}^n$, then show that $W^{\perp}$ a subspace of $\mathbb{R}^n$.
\item Show that if $x$ is in both $W$ and $W^{\perp}$, then $x=0$.
\end{enumerate}
\subsection{Orthogonal Sets}
A set of vectors $\{ u_1, ...u_p\}$ in a vector space $V$ is said to be \textbf{orthogonal set} if each pair of distinct vectors from the set is orthogonal,i.e for all $u_i, u_j \in V$ we have $u_i.u_j = 0$ whenever $i \neq j$.

@@

\subsection{Exercise}
\begin{enumerate}
\item Determine whether the following sets of vectors are orthogonal or not.
  \begin{enumerate}
  \item $
    \begin{bmatrix}
      1 \\ -2 \\ 1
    \end{bmatrix}, \vspace{5mm}
    \begin{bmatrix}
      0 \\ 1 \\ 2
    \end{bmatrix},  \vspace{5mm}
    \begin{bmatrix}
      -5 \\ -2 \\ 1
    \end{bmatrix}$
  \end{enumerate}

\item Show that $\{u_1, u_2\}$ is an orthogonal basis of $\mathbb{R}^2$ and find the coordinates of $x$ in terms of this basis.
  \begin{enumerate}
  \item   $u_1 =
  \begin{bmatrix}
    2 \\ 3
  \end{bmatrix} ,  \vspace{5mm}
  u_2 =
  \begin{bmatrix}
    6 \\ 4
  \end{bmatrix},  \vspace{5mm}
  x =
  \begin{bmatrix}
    9 \\ -7
  \end{bmatrix}
  $

 \item $u_1 =
  \begin{bmatrix}
    3 \\ 1
  \end{bmatrix} ,  \vspace{5mm}
  u_2 =
  \begin{bmatrix}
    2 \\ -6
  \end{bmatrix},  \vspace{5mm}
  x =
  \begin{bmatrix}
    -6 \\ 3
  \end{bmatrix}
  $

  \item Compute the orthogonal projection of $
    \begin{bmatrix}
      1 \\ 7
    \end{bmatrix}
    $ onto to the line through $
    \begin{bmatrix}
      -4 \\ 2
    \end{bmatrix}
    $ and the origin.

  \item Compute the distance of $ y
    \begin{bmatrix}
      3 \\ 1
    \end{bmatrix}
    $ to the line through
    $u= \begin{bmatrix}
      8 \\ 6
    \end{bmatrix}
    $ and the origin.
    \end{enumerate}
\end{enumerate}
\subsection{Orthogonal Projection}
The coordinate $c_j$ of $y$ in Theorem~\ref{orthopro} is actually orthogonal projection of $y$ into the vector $u_j$. This can be generalized. For any given vector $u$. The orthogonal projection of a vector $y$ on $u$ is given by the formula $\displaystyle \hat{y} = \frac{y.u}{u.u}$

### 2025-07-03 lecturenotes_aima203/ch6_aima203.tex
orthogonal

\begin{remark}
  It should be noted that this rule requires the division of the whole range into an even number of sub-intervals of width $h$.
\end{remark}

\subsection{Simpson's 3/8-Rule}
Setting $n=3$ in the general formula ~\ref{numint}, all differences higher than the third will become zero and we obtained \\[1mm]
$\displaystyle \int_{x_0}^{x_3} \; y\,dx = 3h \left [ y_0 + \frac{3}{2} \Delta \, y_0 + \frac{3}{4} \Delta^2 \, y_0 + \frac{1}{8} \Delta ^3 \, y_0 \right ] = \frac{3h}{8}\, (y_0 + 3y_1 + 3y_2+y_3).$ \\ Similarly we can obtain \\[5mm] $\int_{x_3}^{x_6} \; y\,dx, ..... \int_{x_{n-3}}^n \; y\,dx $. Then we have,
\begin{equation}
  \label{simp2}
  \begin{aligned}[b]
    \int_{x_0}^{x_n}\; y\,dx &= \int_{x_0}^{x_3} \; y\,dx + \int_{x_3}^{x_6} \; y\,dx + .... + \int_{x_{n-3}}^{x_n} \; y\,dx \\[1mm]
                             &= \frac{3h}{8} \, (y_0 +3y_1 + 3y_2 + y_3) + \frac{3h}{8} \, (y_3 + 3y_4 + 3y_5 + y_6) + .... + \frac{3h}{8} \, (y_{n-3} + 3y_{n-2} + 3y_{n-1} + y_n) \\[1mm]
                             &= \frac{3h}{8} \, \{ y_0 + 3(y_1 + y_2 + y_4 + y_5 + ... + y_{n-1}) + 2 (y_3 + y_6  ... + y_{n-3} ) + y_n \}
  \end{aligned}
\end{equation}

\subsection{Exercise}
\begin{enumerate}
\item For the given $(x,y)$ data points: Evaluate $\int_0^2 \, ydx$ using the trapezoidal rule.\\  $(0.0,0.399),\, (0.5,0.352),\,(1.0,0.242),\, (1.5,0.129),\,(2.0,0.054)$.

@@

\begin{figure}[h]
  \centering
\includegraphics[width=12cm, height=9cm]{gaussian.png}
\end{figure}
\subsection{Exercise}
\begin{enumerate}
 \item Compute the values of $I = \displaystyle \int_0^1 \, \frac{dx}{1+x^2}$ using the trapezoidal rule with $h=0.5, \; 0.25, \; 0.125$. Then obtain a better estimate using \textbf{Romberg's method}.

\item Compute $\displaystyle I\left (\frac{\pi}{4} \right ), \; I\left (\frac{\pi}{8} \right ), \; I\left (\frac{\pi}{8}, \frac{\pi}{4} \right )$ for : $I = \displaystyle \int_0^{\pi/4} \, cos^2x\, dx$.
\item Derive the Gauss quadrature formula for $n=2$ and apply it to evaluate the integral $\displaystyle \int_{-1}^1 \, \frac{dx}{1+x^2}$.

>>>>

\begin{remark}
  It should be noted that this rule requires the division of the whole range into a number of sub-intervals that is a multiple of 2, of width $h$.
\end{remark}

\subsection{Simpson's 3/8-Rule}
Setting $n=3$ in the general formula ~\ref{numint}, all differences higher than the third will become zero and we obtained \\[1mm]
$\displaystyle \int_{x_0}^{x_3} \; y\,dx = 3h \left [ y_0 + \frac{3}{2} \Delta \, y_0 + \frac{3}{4} \Delta^2 \, y_0 + \frac{1}{8} \Delta ^3 \, y_0 \right ] = \frac{3h}{8}\, (y_0 + 3y_1 + 3y_2+y_3).$ \\ Similarly we can obtain \\[5mm] $\int_{x_3}^{x_6} \; y\,dx, ..... \int_{x_{n-3}}^n \; y\,dx $. Then we have,
\begin{equation}
  \label{simp2}
  \begin{aligned}[b]
    \int_{x_0}^{x_n}\; y\,dx &= \int_{x_0}^{x_3} \; y\,dx + \int_{x_3}^{x_6} \; y\,dx + .... + \int_{x_{n-3}}^{x_n} \; y\,dx \\[1mm]
                             &= \frac{3h}{8} \, (y_0 +3y_1 + 3y_2 + y_3) + \frac{3h}{8} \, (y_3 + 3y_4 + 3y_5 + y_6) + .... + \frac{3h}{8} \, (y_{n-3} + 3y_{n-2} + 3y_{n-1} + y_n) \\[1mm]
                             &= \frac{3h}{8} \, \{ y_0 + 3(y_1 + y_2 + y_4 + y_5 + ... + y_{n-1}) + 2 (y_3 + y_6  ... + y_{n-3} ) + y_n \}
  \end{aligned}
\end{equation}
\begin{remark}
  It should be noted that this rule requires the division of the whole range into a number of sub-intervals that is a multiple of 3, of width $h$.
\end{remark}
\subsection{Exercise}
\begin{enumerate}
\item For the given $(x,y)$ data points: Evaluate $\int_0^2 \, ydx$ using the trapezoidal rule.\\  $(0.0,0.399),\, (0.5,0.352),\,(1.0,0.242),\, (1.5,0.129),\,(2.0,0.054)$.

@@

\begin{figure}[h]
  \centering
\includegraphics[width=12cm, height=9cm]{gaussian.png}
\end{figure}
\subsection{Exercise}
\begin{enumerate}
 \item Compute the values of $I = \displaystyle \int_0^1 \, \frac{dx}{1+x^2}$ using the trapezoidal rule with $h=0.5, \; 0.25, \; 0.125$. Then obtain a better estimates using \textbf{Romberg's method}.

\item Compute the values of $I = \displaystyle \int_0^{\pi/2} \, cos^2x\, dx$ using the trapezoidal rule with $h= \pi /4, \; \pi/8$. Then obtain a better estimates using \textbf{Romberg's method}.
\item Derive the Gauss quadrature formula for $n=2$ and apply it to evaluate the integral $\displaystyle \int_{-1}^1 \, \frac{dx}{1+x^2}$.
